---
title: "Research Analysis Notebook"
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

This quarto doc is used to analyze the data.

```{r}
#| label: loading-data
load(here("data", "data_constructed", "analytical_data.RData"))
```

# Research Question:
- How does discrimination affect depression risk?

# Data Exploration (?)
- 6% of the data shows positive phq2 screener (and there for at 
risk for depression)
- for sexual orientation 95% of the respondents identified themselves as straight, while less than two percent of the population identified with the each of the remaining options
- the proportion of non-straight sexual orientations are larger in the depressed population
- Interaction of all three of sex, race, sexual_orientation leads to singleton 

```{r}
#| label: prop-tables

prop.table(table(final_nhis$phq2_screener_result))

prop.table(table(final_nhis$sexual_orientation))

depressed_sample <- final_nhis|>
  filter(phq2_screener_result == "Positive")

prop.table(table(depressed_sample$sexual_orientation))

table(final_nhis$phq2_screener_result, final_nhis$sex, final_nhis$hisprace, final_nhis$sexual_orientation)

# table(depressed_sample$sex, depressed_sample$sexual_orientation, depressed_sample$hisprace)
```
# Some visuals (exploratory)
- summated scale is a exponential distribution for both the full sample and the individuals with positive depression risk
- positive depression risk seems to have a higher intercept, but realitive parallel coefficients which indicate an increase in caution is related to an increase on prejudice.



```{r}
#| label: some-visualizations

# Histogram of distribution of summated with fill as phq2 screener result
ggplot(final_nhis, aes(x = summated, fill = phq2_screener_result))+
  geom_histogram()+
  theme_bw()

ggplot(depressed_sample, aes(x = summated, fill = sexual_orientation))+
  geom_histogram()+
  theme_bw()

ggplot(final_nhis, aes(x = summated, y = phq2_screener_result))+
  geom_jitter()+
  geom_point()+
  theme_bw()

ggplot(final_nhis, aes(x = dissmart, y = phq2_screener_result))+
  geom_jitter()+
  geom_point()+
  theme_bw()

ggplot(final_nhis, aes(x = vigavoid, y = phq2_screener_result))+
  facet_wrap(~hisprace)+
  geom_jitter()+
  geom_point()+
  theme_bw()

ggplot(final_nhis, aes(x = summated, y = phq2_screener_result, color = sexual_orientation))+
  facet_wrap(~hisprace)+
  geom_jitter()

ggplot(final_nhis, aes(x = caution_scale, y = prejudice_scale, color = phq2_screener_result))+
  geom_point()+
  geom_smooth()+
  theme_bw()
```
```{r}
#| label: co-linearity-visuals

numeric <- final_nhis |>
  dplyr::select(-age)|>
  mutate(
    disrespct = as.numeric(disrespct),
    disserv = as.numeric(disserv),
    disfear = as.numeric(disfear),
    disthreat = as.numeric(disthreat),
    viginsult = as.numeric(viginsult),
    vigappear = as.numeric(vigappear),
    vigsay = as.numeric(vigsay),
    vigavoid = as.numeric(vigavoid),
  ) |>
  select_if(is.numeric)

corrgram(numeric, upper.panel="panel.cor", order = TRUE)
```
# Models

```{r}
#| label: logit_model_without_survey_design_stuff_prejudice_scale

model <- glm(phq2_screener_result ~ prejudice_scale, data = final_nhis, family = binomial(logit))
model1 <- glm(phq2_screener_result ~ prejudice_scale + sex + hisprace + sexual_orientation, data = final_nhis, family = binomial(logit))
# model2 <- glm(phq2_screener_result ~ prejudice_scale*sex*hisprace*sexual_orientation, data = final_nhis, family = binomial(logit))
modelsummary(list(model, model1), stars = TRUE)
```

```{r}
#| label: logit_model_without_survey_design_stuff_summated

model <- glm(phq2_screener_result ~ summated, data = final_nhis, family = binomial(logit))
# model1 <- glm(phq2_screener_result ~ summated * hisprace, data = final_nhis, family = binomial(logit))
model2 <- glm(phq2_screener_result ~ summated + sex + hisprace + sexual_orientation, data = final_nhis, family = binomial(logit))
# model3 <- glm(phq2_screener_result ~ summated*sex*hisprace*sexual_orientation, data = final_nhis, family = binomial(logit))
modelsummary(list(model, model2), stars = TRUE)
```
# THIS ONE 
For interpreting scales, one standard deviation of change in a scale is associated with change in odds
- TODO: explore marginal effects and predicts
-The first model is representative of log odds by demographics (without interaction terms)
- References: sexMale, non-hispanic white, straight
-The second model takes into consideration a scale (TODO: rename prejudice_summated to something better)
  -Interpretation: For one standard deviation increase in the predjudice scale...

Insights
- Change of effect when taking into consideration of discrimination, vigilience, and summated scales
  - sexFemale affect peaks when taking into consideration the prejudice scale at exp(.107) odds greater than men
  - when accounting for the scales, the affect of Black disappates
  - some of the affect of bisexualiy and other sexualities is accounted for by discrimintaion but not all
  - the affects of accounting for scales has no impact of the odds a person who identified with Asian, and in general they are less likely to have a positive depression screener result

```{r}
#| label: BIC-comparisons-scales
model.demographics <- glm(phq2_screener_result ~ sex + sexual_orientation + hisprace + I(age-18) + (I(age-18))^2, 
                          data = final_nhis, family = binomial(logit))
model.discrimination <- update(model.demographics, .~. + prejudice_scale)
model.vigilence <- update(model.demographics, .~. + caution_scale)
model.summated <- update(model.demographics, .~. + summated)

modelsummary(list(model.demographics, model.discrimination, model.vigilence, model.summated),
             stars = TRUE)

```


```{r}
#| label: fig-marginal-effects-by-race

predict_baseline <- avg_predictions(model.demographics, variables = "hisprace") |>
  as_tibble() |>
  mutate(model = "baseline")

predict_discrimination <- avg_predictions(model.discrimination, variables = "hisprace") |>
  as_tibble() |>
  mutate(model = "discrimination control")

predict_vigilence <- avg_predictions(model.vigilence, variables = "hisprace") |>
  as_tibble() |>
  mutate(model = "vigilence control")

predict_summated <- avg_predictions(model.summated, variables = "hisprace")|>
  as_tibble() |>
  mutate(model = "summated control")

predict_baseline |>
  bind_rows(predict_discrimination) |>
  # bind_rows(predict_vigilence)|>
  bind_rows(predict_summated) |>
  ggplot(aes(x = hisprace, y = estimate, color = model))+
  geom_point()+
  coord_flip()

```
```{r}
#| label: fig-marginal-effects-by-sex

predict_baseline <- avg_predictions(model.demographics, variables = "sex") |>
  as_tibble() |>
  mutate(model = "baseline")

predict_discrimination <- avg_predictions(model.discrimination, variables = "sex") |>
  as_tibble() |>
  mutate(model = "discrimination control")

predict_vigilence <- avg_predictions(model.vigilence, variables = "sex") |>
  as_tibble() |>
  mutate(model = "vigilence control")

predict_summated <- avg_predictions(model.summated, variables = "sex")|>
  as_tibble() |>
  mutate(model = "summated control")

predict_baseline |>
  bind_rows(predict_discrimination) |>
  bind_rows(predict_vigilence)|>
  bind_rows(predict_summated) |>
  ggplot(aes(x = sex, y = estimate, color = model))+
  geom_point()+
  coord_flip()

```

```{r}
#| label: fig-marginal-effects-by-sexual-orientation

predict_baseline <- avg_predictions(model.demographics, variables = "sexual_orientation") |>
  as_tibble() |>
  mutate(model = "baseline")

predict_discrimination <- avg_predictions(model.discrimination, variables = "sexual_orientation") |>
  as_tibble() |>
  mutate(model = "discrimination control")

predict_vigilence <- avg_predictions(model.vigilence, variables = "sexual_orientation") |>
  as_tibble() |>
  mutate(model = "vigilence control")

predict_summated <- avg_predictions(model.summated, variables = "sexual_orientation")|>
  as_tibble() |>
  mutate(model = "summated control")

predict_baseline |>
  bind_rows(predict_discrimination) |>
  bind_rows(predict_vigilence)|>
  bind_rows(predict_summated) |>
  ggplot(aes(x = sexual_orientation, y = estimate, color = model))+
  geom_point()+
  coord_flip()

```

# Poly - models
Could be useful for information on scoring on specific quesions by demographic information

```{r}
#| label: polytomous-models-cummulative-disrespect

model <- polr(disrespct ~ disrespct, data = final_nhis)
model1 <- update(model, .~. + sexual_orientation)

modelsummary(list(model, model1), stars = TRUE)
```

```{r}
#| label: polytoumous-models-multinomial-sexual-orientation-nominal
#| include: FALSE

model <- multinom(sexual_orientation ~ summated, data = final_nhis)
model1 <- update(model, .~. + hisprace)
model2 <- update(model1, .~. + sex + sexual_orientation)

modelsummary(list(model, model1, model2),
             shape = term + statistic ~ model + response,
             stars = TRUE)

```


# 
```{r}
#| label: poly-models-

model <- polr(disthreat ~ phq2_screener_result, data = final_nhis)
model1 <- update(model, .~. + sex + sexual_orientation + hisprace)

modelsummary(list(model,model1), stars = TRUE)
```
