---
title: "Research Analysis Notebook"
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

This quarto doc is used to analyze the data.

```{r}
#| label: loading-data
load(here("data", "data_constructed", "analytical_data.RData"))
```

# Research Question:
- How does discrimination affect depression risk?

# Data Exploration (?)
- 6% of the data shows positive phq2 screener (and there for at 
risk for depression)
- for sexual orientation 95% of the respondents identified themselves as straight, while less than two percent of the population identified with the each of the remaining options
- the proportion of non-straight sexual orientations are larger in the depressed population
- Interaction of all three of sex, race, sexual_orientation leads to singleton 

```{r}
#| label: prop-tables

prop.table(table(final_nhis$phq2_screener_result))

prop.table(table(final_nhis$sexual_orientation))

depressed_sample <- final_nhis|>
  filter(phq2_screener_result == "Positive")

prop.table(table(depressed_sample$sexual_orientation))

table(final_nhis$phq2_screener_result, final_nhis$sex, final_nhis$hisprace, final_nhis$sexual_orientation)

# table(depressed_sample$sex, depressed_sample$sexual_orientation, depressed_sample$hisprace)
```
# Some visuals (exploratory)
- summated scale is a exponential distribution for both the full sample and the individuals with positive depression risk
- positive depression risk seems to have a higher intercept, but realitive parallel coefficients which indicate an increase in caution is related to an increase on prejudice.



```{r}
#| label: some-visualizations

# Histogram of distribution of summated with fill as phq2 screener result
ggplot(final_nhis, aes(x = summated, fill = phq2_screener_result))+
  geom_histogram()+
  theme_bw()

ggplot(depressed_sample, aes(x = summated, fill = sexual_orientation))+
  geom_histogram()+
  theme_bw()

ggplot(final_nhis, aes(x = summated, y = phq2_screener_result))+
  geom_jitter()+
  geom_point()+
  theme_bw()

ggplot(final_nhis, aes(x = dissmart, y = phq2_screener_result))+
  geom_jitter()+
  geom_point()+
  theme_bw()

ggplot(final_nhis, aes(x = vigavoid, y = phq2_screener_result))+
  facet_wrap(~hisprace)+
  geom_jitter()+
  geom_point()+
  theme_bw()

ggplot(final_nhis, aes(x = summated, y = phq2_screener_result, color = sexual_orientation))+
  facet_wrap(~hisprace)+
  geom_jitter()

ggplot(final_nhis, aes(x = caution_scale, y = prejudice_scale, color = phq2_screener_result))+
  geom_point()+
  geom_smooth()+
  theme_bw()
```
```{r}
#| label: co-linearity-visuals

numeric <- final_nhis |>
  mutate(
    disrespct = as.numeric(disrespct),
    disserv = as.numeric(disserv),
    disfear = as.numeric(disfear),
    disthreat = as.numeric(disthreat),
    viginsult = as.numeric(viginsult),
    vigappear = as.numeric(vigappear),
    vigsay = as.numeric(vigsay),
    vigavoid = as.numeric(vigavoid),
  ) |>
  select_if(is.numeric)

corrgram(numeric, upper.panel="panel.cor", order = TRUE)
```
# Models

```{r}
#| label: logit_model_without_survey_design_stuff_prejudice_scale

model <- glm(phq2_screener_result ~ prejudice_scale, data = final_nhis, family = binomial(logit))
model1 <- glm(phq2_screener_result ~ prejudice_scale + sex + hisprace + sexual_orientation, data = final_nhis, family = binomial(logit))
# model2 <- glm(phq2_screener_result ~ prejudice_scale*sex*hisprace*sexual_orientation, data = final_nhis, family = binomial(logit))
modelsummary(list(model, model1), stars = TRUE)
```

```{r}
#| label: logit_model_without_survey_design_stuff_summated

model <- glm(phq2_screener_result ~ summated, data = final_nhis, family = binomial(logit))
# model1 <- glm(phq2_screener_result ~ summated * hisprace, data = final_nhis, family = binomial(logit))
model2 <- glm(phq2_screener_result ~ summated + sex + hisprace + sexual_orientation, data = final_nhis, family = binomial(logit))
# model3 <- glm(phq2_screener_result ~ summated*sex*hisprace*sexual_orientation, data = final_nhis, family = binomial(logit))
modelsummary(list(model, model2), stars = TRUE)
```
# THIS ONE I THINK
For interpreting scales, one standard deviation of change in a scale is associated with change in odds
- TODO: explore marginal effects and predicts

```{r}
#| label: BIC-comparisons-scales
model.demographics <- glm(phq2_screener_result ~ sex + sexual_orientation + hisprace, 
                          data = final_nhis, family = binomial(logit))
model.discrimination <- update(model.demographics, .~. + prejudice_scale)
model.vigilence <- update(model.demographics, .~. + caution_scale)
model.summated <- update(model.demographics, .~. + summated)

modelsummary(list(model.demographics, model.discrimination, model.vigilence, model.summated),
             stars = TRUE)

```


```{r}
library(marginaleffects)
predict_baseline <- avg_predictions(model.demographics, variables = "hisprace") |>
  as_tibble() |>
  mutate(model = "baseline")

predict_discrimination <- avg_predictions(model.discrimination, variables = "hisprace") |>
  as_tibble() |>
  mutate(model = "discrimination control")

predict_baseline |>
  bind_rows(predict_discrimination) |>
  ggplot(aes(x = hisprace, y = estimate, color = model))+
  geom_point()+
  coord_flip()

```

```{r}
#| label: polytomous-models-cummulative-disrespect

model <- polr(disrespct ~ sex, data = final_nhis)
model1 <- update(model, .~. + sexual_orientation)

modelsummary(list(model, model1), stars = TRUE)
```

```{r}
#| label: polytoumous-models-multinomial-sexual-orientation-nominal
#| include: FALSE

model <- multinom(sexual_orientation ~ summated, data = final_nhis)
model1 <- update(model, .~. + hisprace)
model2 <- update(model1, .~. + sex + sexual_orientation)

modelsummary(list(model, model1, model2),
             shape = term + statistic ~ model + response,
             stars = TRUE)

```


# 
```{r}
#| label: poly-models-

model <- polr(disthreat ~ phq2_screener_result, data = final_nhis)
model1 <- update(model, .~. + sex + sexual_orientation + hisprace)

modelsummary(list(model,model1), stars = TRUE)
```
