---
title: "Organize Data Notebook"
---

```{r}
#| label: setup
#| include: false
library(here)
source(here("utils","check_packages.R"))
source(here("utils","functions.R"))
```

This quarto doc is used to organize the data.

Reading the National Health Interview Survey into R
```{r}
#| label: loading data
#| include: false

nhis <- read_fwf(here("data", "data_raw", "nhis_00002.dat.gz"),
                 col_positions = fwf_cols(year = c(1,4),
                                          serial = c(5,10),
                                          strata = c(11,14),
                                          psu = c(15,17),
                                          nhishid = c(18,31),
                                          # pernum = c(32,33), # not in codebook
                                          nishpid = c(34,49),
                                          # hhx = c(50,56), # all NA, not in codebook
                                          sampweight = c(57,68),
                                          astatflg = 69,
                                          cstatflg = 70,
                                          age = c(71,73),
                                          sex = 74,
                                          sexorien = 75,
                                          # marstcur = 76,
                                          hisprace = c(77,78),
                                          phqscreen = 79,
                                          disrespct = 80,
                                          disserv = 81,
                                          dissmart = 82,
                                          disfear = 83,
                                          disthreat = 84,
                                          viginsult = 85,
                                          vigappear = 86,
                                          vigsay = 87,
                                          vigavoid = 88),
                 col_types = c(nhishid = 'c',
                               nishpid = 'c',
                               .default = 'i'),
                 show_col_types = FALSE)
  
```

Cleaning data 

This tidy dataset contains...
   -dependent variables:
   -key independent variables:
   -a few important covariates: 
   
Note: NIU means "Not in Universe" and stuctural skip
   
```{r}
#| label: clean data
#| include: true

nhis <-nhis|>
  mutate(sample_flag = case_when(astatflg == 0 & cstatflg == 0 ~ NA, # NIU
                                 astatflg == 0 & cstatflg == 1 ~ "Sample child-has record",
                                 astatflg == 0 & cstatflg == 2 ~ 'Sample child-no record',
                                 astatflg == 0 & cstatflg == 3 ~ "Not selected as sample child",
                                 astatflg == 0 & cstatflg == 4 ~ "No one selected as sample child",
                                 astatflg == 0 & cstatflg == 5 ~ "Emancipated minor",
                                 cstatflg == 0 & astatflg == 1 ~ "Sample adult, has record",
                                 cstatflg == 0 & astatflg == 2 ~ "Sample adult, no record",
                                 cstatflg == 0 & astatflg == 3 ~ "Not selected as sample adult",
                                 cstatflg == 0 & astatflg == 4 ~ "No one selected as sample adult",
                                 cstatflg == 0 & astatflg == 5 ~ "Armed forces member",
                                 cstatflg == 0 & astatflg == 6 ~ "AF member, selected as sample adult"),
         sex = case_when(sex == 1 ~ "male",
                         sex == 2 ~ "female",
                         sex == 7 ~ "unknown-refused",
                         sex == 8 ~ "unknown-not ascertained",
                         sex == 9 ~ "unknown-don't know"),
         sexual_orientation = case_when(sexorien == 0 ~  NA, # "NIU"
                                        sexorien == 1 ~ "Lesbian or gay",
                                        sexorien == 2 ~ "Straight, that is, not lesbian or gay",
                                        sexorien == 3 ~ "Bisexual",
                                        sexorien == 4 ~ "Something else",
                                        sexorien == 5 ~ "I don't know the answer",
                                        sexorien == 7 ~ "Unknown-refused",
                                        sexorien == 8 ~ "Unknown-not ascertained"),
         hisprace = case_when(hisprace == 01 ~ "Hispanic",
                              hisprace == 02 ~ "Non-Hispanic White",
                              hisprace == 03 ~ "Non-Hispanic Black/African American",
                              hisprace == 04 ~ "Non-Hispanic Asian",
                              hisprace == 05 ~ "Non-Hispanic American Indian/Alaskan Native",
                              hisprace == 06 ~ "Non-Hispanic American Indian/Alaskan Native and any other group",
                              hisprace == 07 ~ "Non-Hispanic Other single and multiple races",
                              hisprace == 97 ~ "Unknown-refused",
                              hisprace == 98 ~ "Unknown-not ascertained",
                              hisprace == 99 ~ "Unknown-don't know"),
         phq2_screener_result = case_when(phqscreen == 1 ~ "Positive",
                                          phqscreen == 2 ~ "Negative",
                                          phqscreen == 8 ~ "Not ascertained",
                                          phqscreen == 9 ~ NA #"NIU"
                                          ),
         # Discrimination
         disrespct = case_when(disrespct == 0 ~ NA, # NIU
                               disrespct == 1 ~ "At least once a week",
                               disrespct == 2 ~ "A few times a month",
                               disrespct == 3 ~ "A few times a year",
                               disrespct == 4 ~ "Less than once a year",
                               disrespct == 5 ~ "Never",
                               disrespct == 7 ~ "Unknown - refused",
                               disrespct == 8 ~ "Unknown - not ascertained",
                               disrespct == 9 ~ "Unknown - don't know"),
         disserv = case_when(disserv == 0 ~ NA, # NIU
                             disserv == 1 ~ "At least once a week",
                             disserv == 2 ~ "A few times a month",
                             disserv == 3 ~ "A few times a year",
                             disserv == 4 ~ "Less than once a year",
                             disserv == 5 ~ "Never",
                             disserv == 7 ~ "Unknown - refused",
                             disserv == 8 ~ "Unknown - not ascertained",
                             disserv == 9 ~ "Unknown - don't know"),
         dissmart = case_when(dissmart == 0 ~ NA, # NIU
                              dissmart == 1 ~ "At least once a week",
                              dissmart == 2 ~ "A few times a month",
                              dissmart == 3 ~ "A few times a year",
                              dissmart == 4 ~ "Less than once a year",
                              dissmart == 5 ~ "Never",
                              dissmart == 7 ~ "Unknown - refused",
                              dissmart == 8 ~ "Unknown - not ascertained",
                              dissmart == 9 ~ "Unknown - don't know"),
         disfear = case_when(disfear == 0 ~ NA, # NIU
                             disfear == 1 ~ "At least once a week",
                             disfear == 2 ~ "A few times a month",
                             disfear == 3 ~ "A few times a year",
                             disfear == 4 ~ "Less than once a year",
                             disfear == 5 ~ "Never",
                             disfear == 7 ~ "Unknown - refused",
                             disfear == 8 ~ "Unknown - not ascertained",
                             disfear == 9 ~ "Unknown - don't know"),
         disthreat = case_when(disthreat == 0 ~ NA, # NIU
                               disthreat == 1 ~ "At least once a week",
                               disthreat == 2 ~ "A few times a month",
                               disthreat == 3 ~ "A few times a year",
                               disthreat == 4 ~ "Less than once a year",
                               disthreat == 5 ~ "Never",
                               disthreat == 7 ~ "Unknown - refused",
                               disthreat == 8 ~ "Unknown - not ascertained",
                               disthreat == 9 ~ "Unknown - don't know"),
         # Vigilence 
         viginsult = case_when(viginsult == 0 ~ NA, # NIU
                               viginsult == 1 ~ "At least once a week",
                               viginsult == 2 ~ "A few times a month",
                               viginsult == 3 ~ "A few times a year",
                               viginsult == 4 ~ "Less than once a year",
                               viginsult == 5 ~ "Never",
                               viginsult == 7 ~ "Unknown - refused",
                               viginsult == 8 ~ "Unknown - not ascertained",
                               viginsult == 9 ~ "Unknown - don't know"),         
         vigappear = case_when(vigappear == 0 ~ NA, # NIU
                               vigappear == 1 ~ "At least once a week",
                               vigappear == 2 ~ "A few times a month",
                               vigappear == 3 ~ "A few times a year",
                               vigappear == 4 ~ "Less than once a year",
                               vigappear == 5 ~ "Never",
                               vigappear == 7 ~ "Unknown - refused",
                               vigappear == 8 ~ "Unknown - not ascertained",
                               vigappear == 9 ~ "Unknown - don't know"),
         vigsay = case_when(vigsay == 0 ~ NA, # NIU
                            vigsay == 1 ~ "At least once a week",
                            vigsay == 2 ~ "A few times a month",
                            vigsay == 3 ~ "A few times a year",
                            vigsay == 4 ~ "Less than once a year",
                            vigsay == 5 ~ "Never",
                            vigsay == 7 ~ "Unknown - refused",
                            vigsay == 8 ~ "Unknown - not ascertained",
                            vigsay == 9 ~ "Unknown - don't know"),
         vigavoid = case_when(vigavoid == 0 ~ NA, # NIU
                              vigavoid == 1 ~ "At least once a week",
                              vigavoid == 2 ~ "A few times a month",
                              vigavoid == 3 ~ "A few times a year",
                              vigavoid == 4 ~ "Less than once a year",
                              vigavoid == 5 ~ "Never",
                              vigavoid == 7 ~ "Unknown - refused",
                              vigavoid == 8 ~ "Unknown - not ascertained",
                              vigavoid == 9 ~ "Unknown - don't know")
  )

nhis <- nhis|>
  dplyr::select(c(-year,-astatflg,-cstatflg,-sexorien, -phqscreen)) # need to drop year, astatflg, cstatflg, sexorien, phqscreen,




```

Dealing with Missing Values
- observations with labeled as children are valid skips in data

```{r}
#| label: Deals with valid skips
#| include: true

# include only samples from adults
nhis <- nhis |>
  filter(sample_flag == "Sample adult, has record")
```

Dealing with missingness (other than valid skip)
- for first pass, I exclude all of responses containing unknown reason agnostic

```{r}
#| label: Deals with other types of missingness
#| include: true

# case-wise deletion of not ascertained or refused
nhis[nhis == "Unknown - not ascertained"] <- NA
nhis[nhis == "Unknown - refused"] <- NA
nhis[nhis == "Unknown - don't know"] <- NA
nhis[nhis == "Not ascertained"] <- NA
nhis <- na.omit(nhis)

# dealing with out of range ages
nhis <- nhis |>
  filter(age < 100 & age > 17)

```
  

Transforming variables into factor variables
```{r}
#| label: transforms variables
#| include: true

nhis <- nhis|>
  mutate(sex = factor(x = sex, levels = c("male", "female")),
         sexual_orientation = factor(sexual_orientation, levels = c("Lesbian or gay",
                                                                    "Straight, that is, not lesbian or gay",
                                                                    "Bisexual",
                                                                    "Something else")),
         hisprace = factor(hisprace, levels = c("Hispanic", 
                                                "Non-Hispanic White",
                                                "Non-Hispanic Black/African American",
                                                "Non-Hispanic Asian",
                                                "Non-Hispanic American Indian/Alaskan Native",
                                                "Non-Hispanic American Indian/Alaskan Native and any other group",
                                                "Non-Hispanic Other single and multiple races")),
         phq2_screener_result = factor(phq2_screener_result, levels = c("Negative", "Positive")),
         
         # Discrimination
         disrespct = factor(disrespct, levels = c("Never",
                                                  "Less than once a year",
                                                  "A few times a year",
                                                  "A few times a month",
                                                  "At least once a week")),
         disserv = factor(disserv, levels = c("Never",
                                              "Less than once a year",
                                              "A few times a year",
                                              "A few times a month",
                                              "At least once a week")),
         dissmart = factor(dissmart, levels = c("Never",
                                                "Less than once a year",
                                                "A few times a year",
                                                "A few times a month",
                                                "At least once a week")),
         disfear = factor(disfear, levels = c("Never",
                                            "Less than once a year",
                                            "A few times a year",
                                            "A few times a month",
                                            "At least once a week")),
         disthreat = factor(disthreat, levels = c("Never",
                                                  "Less than once a year",
                                                  "A few times a year",
                                                  "A few times a month",
                                                  "At least once a week")),
         # Vigilence
         viginsult = factor(viginsult, levels = c("Never",
                                                  "Less than once a year",
                                                  "A few times a year",
                                                  "A few times a month",
                                                  "At least once a week")),
         vigappear = factor(vigappear, levels = c("Never",
                                                  "Less than once a year",
                                                  "A few times a year",
                                                  "A few times a month",
                                                  "At least once a week")),
         
         
         vigsay = factor(vigsay, levels = c("Never",
                                            "Less than once a year",
                                            "A few times a year",
                                            "A few times a month",
                                            "At least once a week")),
         vigavoid = factor(vigavoid, levels = c("Never",
                                                "Less than once a year",
                                                "A few times a year",
                                                "A few times a month",
                                                "At least once a week")),
  )


```


Spliting the discrimination and heightened vigilence questions into two factors reduces cronbach's alpha.
```{r}
#| label: factor_loading

numeric <- nhis |>
  select_if(is.factor)|>
  mutate(
    disrespct = as.numeric(disrespct),
    disserv = as.numeric(disserv),
    disfear = as.numeric(disfear),
    disthreat = as.numeric(disthreat),
    viginsult = as.numeric(viginsult),
    vigappear = as.numeric(vigappear),
    vigsay = as.numeric(vigsay),
    vigavoid = as.numeric(vigavoid),
  )|>
  dplyr::select(disrespct, disserv, disfear, disthreat, viginsult, vigappear, vigsay, vigavoid)

numeric_r <- cor(numeric, use = "pairwise.complete.obs")
alpha(numeric_r)

fa1 <- fa(numeric_r, 1)
fa2 <- fa(numeric_r, 2)
# fa3 <- fa(numeric_r, 3)
fa.diagram(fa1)
fa.diagram(fa2)

# based on factor analysis, the variables disrespct, disserv, disfear, disthreat, viginsult load onto one factor
# while the remaining vigilence questions load onto another
factor_1 <- numeric|>
  dplyr::select(-vigappear,-vigsay,-viginsult)
alpha(cor(factor_1))
```
```{r}
#| label: one_factor

nhis$disrespct.z <- scale(as.numeric(nhis$disrespct)) 
nhis$disserv.z <- scale(as.numeric(nhis$disserv))
nhis$dissmart.z <- scale(as.numeric(nhis$dissmart))
nhis$disfear.z <- scale(as.numeric(nhis$disfear))
nhis$disthreat.z <- scale(as.numeric(nhis$disthreat))
nhis$viginsult.z <- scale(as.numeric(nhis$viginsult))
nhis$vigappear.z <- scale(as.numeric(nhis$vigappear))
nhis$vigsay.z <- scale(as.numeric(nhis$vigsay))
nhis$vigavoid.z <- scale(as.numeric(nhis$vigavoid))

nhis$summated <- scale(nhis$disrespct.z+
                         nhis$disserv.z+
                         nhis$dissmart.z+
                         nhis$disfear.z+
                         nhis$disthreat.z+
                         nhis$viginsult.z+
                         nhis$vigappear.z+
                         nhis$vigsay.z+
                         nhis$vigavoid.z)
```

```{r}
#| label: factor-loading-vig-without-vig-insult-summated

nhis$caution_scale <- scale(nhis$vigappear.z+
                             nhis$vigsay.z+
                             nhis$vigavoid.z)
```

```{r}
#| label: factor-loadingdiscrimination-and-viginsult-summated

nhis$prejudice_scale <- scale(nhis$viginsult.z+
                             nhis$disrespct.z+
                             nhis$disserv.z+
                             nhis$dissmart.z+
                             nhis$disfear.z)
```

select data
```{r}
#| label: bare_bones_data
#| include: true

final_nhis <- nhis |>
  dplyr::select(sex, hisprace, sexual_orientation,
         disrespct, disserv, dissmart, disfear, disthreat, 
         viginsult, vigappear, vigsay, vigavoid, 
         summated, caution_scale, prejudice_scale,
         phq2_screener_result)|>
  drop_na()
```

Saving data as RData
```{r}
#| label: saving data as analytical_data.RData
#| include: false

save(final_nhis, file = here("data", "data_constructed", "analytical_data.RData"))

```